A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2015-11-18 LensingPaper_draft_v3.01.tex
Included corrections in abstract and introduction.
More corrections; abstract is now 250 words long; edited introduction.
Edited modelling section. Some figure corrections.

\documentclass[useAMS,usenatbib]{mn2e}

@@

%_____MACROS______
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand*\Diff[1]{\mathop{}\!\mathrm{d^#1}}
%\newcommand{\vect}[1]{\boldsymbol{#1}} % Uncomment for BOLD vectors.
\newcommand{\vect}[1]{\vec{#1}} % Uncomment for ARROW vectors.

@@

\date{Accepted [TO DO]. Received [TO DO]; in original form [TO DO]}

@@

%--------------------------------------------
%Abstract
\begin{abstract}
\input{docs/abstract_v2.01}
\end{abstract}
%--------------------------------------------

\begin{keywords}
blabla -- blabla: bla.
\end{keywords}

%--------------------------------------------
%Introduction
\input{docs/introduction_v2.01}
%--------------------------------------------

%------------------------
\section{Modelling} \label{sec:Modelling}
\input{docs/model_MGE_v2.01}
\input{docs/model_lensing_v2.01}
\input{docs/model_JAM_v2.01}
%------------------------

%------------------------
\input{docs/data_v2.01}
%------------------------

%------------------------
\section{Results} \label{sec:Results}
\input{docs/results_photometry_v2.01}
\input{docs/results_lensing_v2.01}
\input{docs/results_JAMwithSB_v2.01}
\input{docs/results_JAMwithNFW_v2.01}
%------------------------

%------------------------
\section{Discussion and Conclusion} \label{sec:Discussion}
\input{docs/discussion_ML_v2.01}
\input{docs/discussion_merger_v2.01}
\input{docs/summary_v2.01}
%------------------------

%---------------------------------------------------------------------------
\bibliographystyle{mn2e}
\bibliography{literaturelist}
%---------------------------------------------------------------------------

>>>>

\documentclass[useAMS,usenatbib]{mnras}

@@

%_____MACROS______
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand*\Diff[1]{\mathop{}\!\mathrm{d^#1}}
%\newcommand{\vect}[1]{\boldsymbol{#1}} % Uncomment for BOLD vectors.
\newcommand{\vect}[1]{\vec{#1}} % Uncomment for ARROW vectors.
\usepackage[usenames,dvipsnames]{xcolor}
\newcommand{\Wilma}[1]{\textcolor{Magenta}{#1}}

@@

\date{Accepted ???. Received ???; in original form ???}

@@

%--------------------------------------------
%Abstract
\begin{abstract}
\input{docs/abstract_v3.02} %[Fertig.]
\end{abstract}
%--------------------------------------------

\begin{keywords}
gravitational lensing: strong -- stars: kinematics and dynamics -- galaxies: kinematics and dynamics -- galaxies: photometry -- galaxies: structure
\end{keywords}

%--------------------------------------------
%Introduction
\input{docs/introduction_v3.02} %[Fertig gekuerzt. Es fehlen noch 2 Referenzen, Zeug im Plot und Plot-Caption.]
%--------------------------------------------

%------------------------
\input{docs/data_v3.02}
%------------------------

%------------------------
\section{Modelling} \label{sec:Modelling}
\input{docs/model_MGE_v3.02} %[Fertig]
\input{docs/model_lensing_v3.01} %[Fertig]
\input{docs/model_JAM_v3.02} %[Fertig - wenn auch nicht so toll.]
%------------------------

%------------------------
\section{Results} \label{sec:Results}
\input{docs/results_photometry_v3.01}
\input{docs/results_lensing_v3.01}
\input{docs/results_JAMwithSB_v3.01}
\input{docs/results_JAMwithNFW_v3.01}
%------------------------

%------------------------
\section{Discussion and Conclusion} \label{sec:Discussion}
\input{docs/discussion_ML_v3.01}
\input{docs/discussion_merger_v3.01}
\input{docs/summary_v3.01}
%------------------------

%---------------------------------------------------------------------------
\bibliographystyle{mnras}
\bibliography{literaturelist}
%---------------------------------------------------------------------------

### 2015-11-18 archive/v3/docs/introduction_v3.01.tex
Edited modelling section. Some figure corrections.

%Dynamical modelling
The mass profile at larger galactocentric radii can be probed by gas rotation curves that directly measure the galaxy's circlar velocity profile \Wilma{[TO DO: REF]}. However, due to its dissipative nature, gas motions are very sensitive to disturbances by e.g. spiral arms and bars \Wilma{[TO DO: REF]}.

@@

\begin{figure*}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/first_glimpse_450.ps}
  \caption{J1331 in F450W}
  \label{fig:F450W}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/first_glimpse_814.ps}
  \caption{J1331 in F814W}
  \label{fig:F814W}
\end{subfigure}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/lens_imgpos.ps}
  \caption{The lensing images}
  \label{fig:lens_just_imgpos}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/stellar_kinematics_data.ps}
  \caption{Stellar Kinematics by \citet{SWELLSV}}
  \label{fig:kinematics}
\end{subfigure}
\caption{Hubble Space telescope (HST) images and stellar kinematics of the galaxy SDSS J1331+3628 (J1331), which has a large counter-rotating core and whose bulge acts as a strong lens for a bluish background source. \emph{Panel (a) and (b):} HST/WFPC2/WFC3 images of J1331 by \citet{SWELLSI} in two filters, F450W in panel (a) and F814W in panel (b). The galaxy's coordinates on the sky are right ascension $\alpha$ = 202.91800$^\circ$ and declination $\delta$ = 36.46999$^\circ$ (epoch J2000). Image orientation and scaling are indicated in panel (a); the scaling transformation from arcseconds to the physical size of the galaxy in kpc uses the galaxy's redshift $z_d = 0.113$ \citep{SWELLSIII} (i.e. assumes an angular diameter distance of 414 Mpc). The color scaling of these two images is the same. The black solid line in panel (b) shows the orientation of the major-axis. The line has a length of $10''$ and indicates the region within which we carry out the Jeans modelling. \emph{Panel (c):} The central region of J1331 in F450W, surface brightness subtracted. An IRAF \emph{ellipse} fit to the F450W surface brightness in panel (a) was subtracted from the image. The (smoothed) residuals within the white square in panel (a) are shown in panel (c). Four bright blobs (A,B,C and D) become visible, which are arranged in a typical strong lensing configuration around the center of the galaxy (G). \emph{Panel (d):} Stellar Kinematics along the galaxy's major axis as measured by \citet{SWELLSV}, line-of-sight rotation velocity $v_\text{rot}$, line-of-sight velocity dispersion $\sigma$ and the rms-velocity $v_\text{rms} = \sqrt{v_\text{rot}^2 + \sigma^2}$. The dotted line in panel (b) indicates the galaxy's effective half-light radius (in the F814W filter), $R_\text{eff} = 2.6" = 5.2$ kpc. The $v_\text{rot}$ curve reveals that J1331 is counter-rotating within $R_\text{eff}$. \Wilma{[TO DO: Replace arcsec with '' in panel d)]} \Wilma{[TO DO: Mention, that the configuration of the additional two bright blobs do not suggest that they form a lensing doublet. They are probably background star formation regions as well. The blob above A could be stretched due to lensing and have a faint counter image between B and G. Maybe indicate in plot?]} \Wilma{[TO DO: Use $\text{km s}^{-1}$ instead of km/s]}}
\label{fig:specialJ1331}
\end{figure*}

>>>>

%Dynamical modelling
The mass profile at larger galactocentric radii can be probed by gas rotation curves that directly measure the galaxy's circlar velocity profile \Wilma{[TO DO: REF]}. However, due to its dissipative nature, gas motions are very sensitive to disturbances by, e.g., spiral arms and bars \Wilma{[TO DO: REF]}.

@@

\begin{figure*}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/first_glimpse_450.ps}
  \caption{J1331 in F450W}
  \label{fig:F450W}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/first_glimpse_814.ps}
  \caption{J1331 in F814W}
  \label{fig:F814W}
\end{subfigure}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/lens_imgpos.ps}
  \caption{The lensing images}
  \label{fig:lens_just_imgpos}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/stellar_kinematics_data.ps}
  \caption{Stellar Kinematics by \citet{SWELLSV}}
  \label{fig:kinematics}
\end{subfigure}
\caption{Hubble Space telescope (HST) images and stellar kinematics of the galaxy SDSS J1331+3628 (J1331), which has a large counter-rotating core and whose bulge acts as a strong lens for a bluish background source. \emph{Panel (a) and (b):} HST/WFPC2/WFC3 images of J1331 by \citet{SWELLSI} in two filters, F450W in panel (a) and F814W in panel (b). The galaxy's coordinates on the sky are right ascension $\alpha$ = 202.91800$^\circ$ and declination $\delta$ = 36.46999$^\circ$ (epoch J2000). Image orientation and scaling are indicated in panel (a); the scaling transformation from arcseconds to the physical size of the galaxy in kpc uses the galaxy's redshift $z_d = 0.113$ \citep{SWELLSIII} (i.e., assumes an angular diameter distance of 414 Mpc). The color scaling of these two images is the same. The black solid line in panel (b) shows the orientation of the major-axis. The line has a length of $10''$ and indicates the region within which we carry out the Jeans modelling. \emph{Panel (c):} The central region of J1331 in F450W, surface brightness subtracted. An IRAF \emph{ellipse} fit to the F450W surface brightness in panel (a) was subtracted from the image. The (smoothed) residuals within the white square in panel (a) are shown in panel (c). Four bright blobs (A,B,C and D) become visible, which are arranged in a typical strong lensing configuration around the center of the galaxy (G). \emph{Panel (d):} Stellar Kinematics along the galaxy's major axis as measured by \citet{SWELLSV}, line-of-sight rotation velocity $v_\text{rot}$, line-of-sight velocity dispersion $\sigma$ and the rms-velocity $v_\text{rms} = \sqrt{v_\text{rot}^2 + \sigma^2}$. The dotted line in panel (b) indicates the galaxy's effective half-light radius (in the F814W filter), $R_\text{eff} = 2.6" = 5.2$ kpc. The $v_\text{rot}$ curve reveals that J1331 is counter-rotating within $R_\text{eff}$. \Wilma{[TO DO: Replace arcsec with '' in panel d)]} \Wilma{[TO DO: Mention, that the configuration of the additional two bright blobs do not suggest that they form a lensing doublet. They are probably background star formation regions as well. The blob above A could be stretched due to lensing and have a faint counter image between B and G. Maybe indicate in plot?]} \Wilma{[TO DO: Use $\text{km s}^{-1}$ instead of km/s]}}
\label{fig:specialJ1331}
\end{figure*}

### 2015-11-18 archive/v3/docs/model_MGE_v3.01.tex
Edited modelling section. Some figure corrections.

\paragraph{Multi-Gaussian Expansion (MGE).} MGEs are used to parametrize the observed surface brightness or projected total mass of a galaxy as a sum of $N$ two-dimensional, elliptical Gaussians \citep{1991ApJ...366..599B,1992A&A...253..366M,1994A&A...285..723E,1999MNRAS.303..495E}. This work makes use of the algorithm and code\footnote{The IDL code package for fitting MGEs to images by \citet{Cap02} is available online at \url{http://www-astro.physics.ox.ac.uk/~mxc/software}. The version from June 2012 was used in this work.} by \citet{Cap02}. We assume all Gaussians to have the same center and position angle $\phi$, i.e. orientation of the Gaussians' major axis measured counter-clockwise from the $y'$-axis of the coordinate system with polar coordinates $(R',\theta')$. Then the surface brightness can be written as
\begin{eqnarray}
I(R',\theta') &=& \sum_{i=1}^{N} I_{0,i} \exp\left[ - \frac{1}{2\sigma_i^2} \left(x^{'2} + \frac{y^{'2}}{q_i^{'2}}\right)\right]\label{eq:MGEgeneral}\\
\text{with } I_{0,i} &=& \frac{L_i}{2\pi \sigma_i^2 q'_i}\label{eq:centralItotalL}\\
\text{and } x'_i &=& R' \cos(\theta' - \phi)\nonumber\\
y'_i &=& R' \sin(\theta' - \phi),\nonumber
\end{eqnarray}
where $I_{0,i}$ is the central surface brightness of each Gaussian, $L_i$ its total luminosity, $\sigma_i$ its dispersion along the major axis and $q'_i$ the axis ratio between the elliptical Gaussians major and minor axis.

>>>>

\paragraph{Multi-Gaussian Expansion (MGE).} MGEs are used to parametrize the observed surface brightness (or projected total mass) of a galaxy as a sum of $N$ two-dimensional, elliptical Gaussians \citep{1991ApJ...366..599B,1992A&A...253..366M,1994A&A...285..723E,1999MNRAS.303..495E}. This work makes use of the algorithm and code\footnote{The IDL code package for fitting MGEs to images by \citet{Cap02} is available online at \url{http://www-astro.physics.ox.ac.uk/~mxc/software}. The version from June 2012 was used in this work.} by \citet{Cap02}. We assume all Gaussians to have the same center and position angle $\phi$, i.e., the orientation of the Gaussians' major axis measured counter-clockwise from the $y'$-axis of the coordinate system with polar coordinates $(R',\theta')$. Then the surface brightness can be written as
\begin{eqnarray}
I(R',\theta') &=& \sum_{i=1}^{N} I_{0,i} \exp\left[ - \frac{1}{2\sigma_i^2} \left(x^{'2} + \frac{y^{'2}}{q_i^{'2}}\right)\right]\label{eq:MGEgeneral}\\
\text{with } I_{0,i} &=& \frac{L_i}{2\pi \sigma_i^2 q'_i}\label{eq:centralItotalL}\\
\text{and } x'_i &=& R' \cos(\theta' - \phi)\nonumber\\
y'_i &=& R' \sin(\theta' - \phi),\nonumber
\end{eqnarray}
where $I_{0,i}$ is the central surface brightness of each Gaussian, $L_i$ its total luminosity, $\sigma_i$ its dispersion along the major axis and $q'_i$ the axis ratio between the elliptical Gaussians major and minor axis.

### 2015-11-18 docs/discussion_ML_v3.01.tex
Edited modelling section. Some figure corrections.

An alternative JAM model was introduced in Section \ref{sec:results_JAM_SB}, where we used the surface brightness distribution without a halo component but together with an increasing mass-to-light ratio. We found that such a model would be perfectly consistent with the lensing results and predict a total mass-to-light ratio in the center of J1331 of  $\Upsilon_\text{I,tot}(R'\sim0) = 2.53$. This is very close to the stellar mass-to-light ratio we got from \citet{SWELLSI} estimates of the bulge's stellar mass using the Chabrier IMF $\Upsilon_\text{I,*}^\text{Chabrier} = 2.5 \pm 0.6$, i.e. less bottom-heavy (more bluish) population.

>>>>

An alternative JAM model was introduced in Section \ref{sec:results_JAM_SB}, where we used the surface brightness distribution without a halo component but together with an increasing mass-to-light ratio. We found that such a model would be perfectly consistent with the lensing results and predict a total mass-to-light ratio in the center of J1331 of  $\Upsilon_\text{I,tot}(R'\sim0) = 2.53$. This is very close to the stellar mass-to-light ratio we got from \citet{SWELLSI} estimates of the bulge's stellar mass using the Chabrier IMF $\Upsilon_\text{I,*}^\text{Chabrier} = 2.5 \pm 0.6$, i.e., less bottom-heavy (more bluish) population.

### 2015-11-18 docs/discussion_merger_v3.01.tex
Edited modelling section. Some figure corrections.

Standard JAM modelling approaches seem not to work for J1331. A JAM model for J1331 would need to allow for stellar mass-to-light ratio gradients within the galaxy, velocity anisotropy and a dark matter halo. Because of degeneracies between stellar and dark mass, and matter distribution and anisotropy profile, such a dynamical model would not lead to very tight constraints on the model parameters.  A search for color gradients in J1331 and/or investigation of absorption line indices could support or contradict the suspicion of the existence of stellar mass-to-light ratio gradients in J133. Subsequently detailed stellar population analyses of the spectra taken along J1331's major axis should be conducted to constrain the mass-to-light ratio reliably.  \Wilma{[TO DO: Include in this discussion the findings from the color profile.]}

@@

High-resolution integral-field spectroscopy could help with this, allow for spatially resolved stellar population analysis and dynamical modelling in two dimensions. \Wilma{[TO DO: Include in this discussion that this could also help test, if the vrms dip is maybe because of twist, i.e. the measurements where not done along the major axis in the disk]} This would lead to much better understanding of J1331's structure and mass distribution and therefore answer questions on how minor mergers might modify spiral galaxies.

>>>>

Standard JAM modelling approaches seem not to work for J1331. A JAM model for J1331 would need to allow for stellar mass-to-light ratio gradients within the galaxy, velocity anisotropy and a dark matter halo. Because of degeneracies between stellar and dark mass, and matter distribution and anisotropy profile, such a dynamical model would not lead to very tight constraints on the model parameters.  A search for colour gradients in J1331 and/or investigation of absorption line indices could support or contradict the suspicion of the existence of stellar mass-to-light ratio gradients in J1331. Subsequently detailed stellar population analyses of the spectra taken along J1331's major axis should be conducted to constrain the mass-to-light ratio reliably.  \Wilma{[TO DO: Include in this discussion the findings from the color profile.]}

@@

High-resolution integral-field spectroscopy could help with this, allow for spatially resolved stellar population analysis and dynamical modelling in two dimensions. \Wilma{[TO DO: Include in this discussion that this could also help test, if the vrms dip is maybe because of twist, i.e., the measurements where not done along the major axis in the disk]} This would lead to much better understanding of J1331's structure and mass distribution and therefore answer questions on how minor mergers might modify spiral galaxies.

### 2015-11-18 docs/introduction_v3.02.tex
Edited modelling section. Some figure corrections.

Cosmological simulations suggest that cold DM forms cuspy halos following a Navarro-Frenk-White (NFW) profile \citep{NFW96}. However, the existence of central DM density cusps in more massive galaxies depends strongly on the stellar mass-to-light ratio (e.g., \citealt{2011MNRAS.416..322D}), while DM dominated dwarf galaxies strongly favour cored DM halos (e.g., \citealt{1994Natur.370..629M,2001ApJ...552L..23D}). This discrepancy, known as the core-cusp problem, might be resolved by galaxy formation processes such as mergers and outflows (e.g., \citealt{2001ApJ...560..636E,2012MNRAS.421.3464P}).

@@

The mass profile at larger galactocentric radii can be probed by gas rotation curves that directly measure the galaxy's circular velocity profile \Wilma{[TO DO: REF]}. However, due to its dissipative nature, gas motions are very sensitive to disturbances by e.g. spiral arms and bars \Wilma{[TO DO: REF]}.

@@

One of the SWELLS galaxies is the massive spiral SDSS J1331+3628, to which we refer as J1331 in the remainder of this work. It has bluish spiral arms and a large reddish bulge (see Figures \ref{fig:F450W} and \ref{fig:F814W}), which is superimposed by quadruplet of extended bluish images approximately at a distance of $1''$ from the galaxy center (see Figure \ref{fig:lens_just_imgpos}). The lensed object might be a star-forming blob of a background galaxy. J1331 stands out of the SWELLS sample because of its large counter-rotating core (see Figure \ref{fig:kinematics}), which might be an indication that J1331 underwent a merger in its recent past.

@@

The goal of this work is to complement the previous work on J1331 by an in-depth analysis of the matter distribution in J1331's inner regions. We use stellar dynamical modelling in addition to lensing constraints, similar to a study of the Einstein Cross by \citet{GlennEC}. We attempt to disentangle the stellar and DM components and test if employed axisymmetric Jeans models work also in the presence of a counter-rotating core. Ideally this work on J1331 could also help understanding how mergers modify the mass distribution of a galaxy.

This paper is organized as follows: Section \ref{sec:Modelling} gives an overview of the modelling techniques used in this work, Section \ref{sec:data} summarizes the data and Section \ref{sec:Results} presents our results on the surface photometry of J1331 using Multi-Gaussian expansions (Section \ref{sec:MGE_results}), constraints from lensing (Section \ref{sec:results_lensing}) and Jeans modelling based on the surface brightness only (Section \ref{sec:results_JAM_SB}) and including a NFW DM halo (Section \ref{sec:results_JAM_NFW}). Finally Section \ref{sec:Discussion} uses the result to discuss J1331's stellar mass-to-light ratio and possible merger history and starting points for future work.

>>>>

Cosmological simulations suggest that cold DM forms cuspy halos following a Navarro-Frenk-White (NFW) profile \citep{NFW96}. However, the existence of central DM density cusps in massive galaxies depends strongly on the stellar mass-to-light ratio (e.g., \citealt{2011MNRAS.416..322D}) and DM dominated dwarf galaxies even favour DM halos with cores (e.g., \citealt{1994Natur.370..629M,2001ApJ...552L..23D}). This discrepancy, known as the core-cusp problem, might be resolved by galaxy formation processes such as mergers and outflows (e.g., \citealt{2001ApJ...560..636E,2012MNRAS.421.3464P}).

@@

The mass profile at larger galactocentric radii can be probed by gas rotation curves that directly measure the galaxy's circular velocity profile \Wilma{[TO DO: REF]}. However, due to its dissipative nature, gas motions are very sensitive to disturbances by e.g., spiral arms and bars \Wilma{[TO DO: REF]}.

@@

One of the SWELLS galaxies is the massive spiral SDSS J1331+3628, to which we refer as J1331 in the remainder of this work. It has bluish spiral arms and a large reddish bulge (see Figures \ref{fig:F450W} and \ref{fig:F814W}), which is superimposed by a quadruplet of extended bluish images approximately at a distance of $1''$ from the galaxy center (see Figure \ref{fig:lens_just_imgpos}). The lensed object might be a star-forming blob of a background galaxy. J1331 stands out of the SWELLS sample because of its large counter-rotating core (see Figure \ref{fig:kinematics}), which might be an indication that J1331 underwent a merger in its recent past.

@@

The goal of this work is to complement the previous work on J1331 by an in-depth analysis of the matter distribution in J1331's inner regions. We use stellar dynamical modelling in addition to lensing constraints, similar to a study of the Einstein Cross by \citet{GlennEC}. We attempt to disentangle the stellar and DM components and test if employed axisymmetric Jeans models work also in the presence of a counter-rotating core. Ideally, this work on J1331 could also help understanding how mergers modify the mass distribution of a galaxy.

This paper is organized as follows: Section \ref{sec:data} summarizes the data, Section \ref{sec:Modelling} gives an overview of the modelling techniques used in this work, and Section \ref{sec:Results} presents our results on the surface photometry of J1331 using Multi-Gaussian expansions (Section \ref{sec:MGE_results}), constraints from lensing (Section \ref{sec:results_lensing}) and Jeans modelling based on the surface brightness only (Section \ref{sec:results_JAM_SB}) and including a NFW DM halo (Section \ref{sec:results_JAM_NFW}). Finally Section \ref{sec:Discussion} uses the result to discuss J1331's stellar mass-to-light ratio, possible merger history and starting points for future work.
